---
title: "Setting a prior"
author: "Thierry Onkelinx"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting a prior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(inlatools)
```

## "iid" random intercept

### Testing `sigma`

`simulate_iid()` generates a number of iid random intercepts from a Gaussian distribution with standard deviation `sigma`.

```{r}
x <- simulate_iid(sigma = 1)
y <- simulate_iid(sigma = 0.1)
```

`plot()` on the simulated random intercepts yields a density with indication of the quantiles.

```{r, fig.show='hold'}
plot(x)
plot(y)
```

Adding `link = "log"` displays the density after back transformation of the random intercepts using the log-link. The random intercept can be interpreted as a relative effect.

```{r, fig.show='hold'}
plot(x, link = "log")
plot(y, link = "log")
```

Adding `link = "logit"` displays the individual random intercepts after back transformation using the loglit-link. Since the influence of the random intercept depends on base proportion, the plot shows for each random intercept the resulting proportion after adding the random intercept to the base proportion.

```{r, fig.show='hold'}
plot(x, link = "logit")
plot(y, link = "logit")
```

### Testing `tau`

One can specify the precision `tau` ($\tau$) instead of the standard deviation `sigma` ($\sigma$). Note that the precision is the inverse of the variance ($\tau = 1 / \sigma ^ 2$).

```{r}
z <- simulate_iid(tau = 10)
plot(z)
```

You must either specify `sigma` or `tau`, the other must remain `NULL` (the default).

```{r error = TRUE}
simulate_iid(sigma = 0.1, tau = 100)
```

## First order random walk

```{r}
x <- simulate_rw(sigma = 0.1, length = 50)
y <- simulate_rw(sigma = 0.01, length = 50)
```

The default plot displays all simulated time series.

```{r, fig.show='hold'}
plot(x)
plot(y)
```

Setting `type = "quantile"` shows, at each time point, the quantiles of the simulated time series.

```{r, fig.show='hold'}
plot(x, type = "quantile")
plot(y, type = "quantile")
```

For each simulated time series, we can calculate the strongest absolute effect at any time point. `type = "divergence"` shows the top 10 simulated time series with the strongest absolute effect. 

```{r, fig.show='hold'}
plot(x, type = "divergence")
plot(y, type = "divergence")
```

`type = "stationary"` is the opposite of `type = "divergence"`: the top 10 simulated time series with the smallest maximum absolute effect at any time point.

```{r, fig.show='hold'}
plot(x, type = "stationary")
plot(y, type = "stationary")
```

`type = "change"` calculates the number of changes in direction of the time series. It will display the top 10 simulated time series with the highest number of changes.

```{r, fig.show='hold'}
plot(x, type = "change")
plot(y, type = "change")
```

Back transformation is available for each `type` of plot. `link = "log"` back transforms from a log-link. The plot shows the relative effect of the random walk.

```{r, fig.show='hold'}
plot(x, type = "quantile", link = "log")
plot(y, type = "quantile", link = "log")
```

The effect of the logit link depends on the base proportion. Therefore `link = "logit"` will create a subplot for 4 different base proportions: 5%, 10%, 25% and 50%.

```{r, fig.show='hold'}
plot(x, type = "quantile", link = "logit")
plot(y, type = "quantile", link = "logit")
```

## Second order random walk

```{r}
x <- simulate_rw(sigma = 0.1, length = 50, order = 2)
y <- simulate_rw(sigma = 0.01, length = 50, order = 2)
```

```{r, fig.show='hold'}
plot(x)
plot(y)
```

```{r, fig.show='hold'}
plot(x, type = "quantile")
plot(y, type = "quantile")
```

```{r, fig.show='hold'}
plot(x, type = "divergence")
plot(y, type = "divergence")
```

```{r, fig.show='hold'}
plot(x, type = "stationary")
plot(y, type = "stationary")
```

```{r, fig.show='hold'}
plot(x, type = "change")
plot(y, type = "change")
```
